---
title: "MATH 216 Homework 1"
author: "Connor McCormick"
output:
  html_document:
    toc: true
    toc_float: true
    collapsed: false
    smooth_scroll: false
---

```{r, echo=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
# For read_csv() command, which is an improved version of base R's read.csv()
library(readr) 
library(lubridate)
library(ggthemes)
library(forcats)


# Load data sets. Note this assumes this .Rmd files is in the same directory as
# the .csv files.
flights <- read_csv("data/flights.csv") %>% 
  mutate(date=as.Date(date))
weather <- read_csv("data/weather.csv") %>% 
  mutate(date=as.Date(date))
planes <- read_csv("data/planes.csv")
airports <- read_csv("data/airports.csv")
states <- read_csv("data/states.csv")
```





## Admistrative:

Please indicate

* Who you collaborated with: Trisha
* Roughly how much time you spent on this HW so far: 9 hours
* The URL of the RPubs published URL [here](http://rpubs.com/connormccormick/HW1).
* What gave you the most trouble: Flight paths
* Any comments you have: I am still somewhat confused on how to use ifelse statements.  I have used if statements a lot in STATA, which work differently, so that might causing me some difficulties.





## Question 1:

Plot a "time series" of the proportion of flights that were delayed by > 30 minutes on each day.  i.e.
 
* the x-axis should be some notion of time
* the y-axis should be the proportion.

Using this plot, indicate describe the
[seasonality](https://en.wikipedia.org/wiki/Seasonality) of when delays over 30
minutes tend to occur.

```{r, echo=FALSE, message=FALSE, fig.width=12, fig.height=6}

flights_per_day <- flights %>% 
  group_by(date) %>% 
  tally() %>% 
  rename(flights_per_day = n)


delays_per_day <- flights %>% 
  filter(!is.na(dep_delay)) %>%
  group_by(date, actual_delay = dep_delay>=30) %>% 
  select(date, actual_delay) %>%
  filter(actual_delay==TRUE) %>%
  tally() %>% 
  rename(delays_per_day = n)


delays_per_day <- left_join(delays_per_day, flights_per_day, by="date") %>%
  mutate(prop = (delays_per_day/flights_per_day))
  

ggplot(delays_per_day, aes(x=date, y=prop)) +
  geom_point(color = "darkblue") +
  geom_smooth(se=FALSE, color = "firebrick") +
  labs(title="Seasonality", x='Date', y='Proportion of Cancelled Flights') +
  coord_cartesian(ylim=c(0, 0.25)) 



```


**Interpretation** 
The delays per day show a seasonal increase around the summer months and the winter holidays. Looking at the trend line, it is clear that there is an increase in delays greater than 30 minutes at these two periods.



## Question 2:

Some people prefer flying on older planes.  Even though they aren't as nice,
they tend to have more room.  Which airlines should these people favor?

```{r, echo=FALSE, fig.width=12, fig.height=6}

planeage <- flights %>%
  left_join(planes, by = "plane") %>%
  filter(!is.na(year)) %>%
  group_by(plane, carrier) %>% 
  summarize(year = mean(year)) 
  
ggplot(planeage, aes(x=fct_reorder(f=carrier, x=year, fun=median, na.rm=TRUE), y=year)) +
  geom_boxplot(fill="aliceblue") +
  geom_jitter(color = "darkblue", alpha = I(.3), size = 1, width = .5) +
  labs(title="Plane Age by Carrier", x='Carrier', y='Year Built') +
  theme_minimal()

```

**Interpretation** 
The two carriers with the oldest set of airplanes is MQ and AA, which are American Eagle and American Airlines.  Though MQ has the oldest planes, there are only two physical planes in the data set. That makes American Airlines the oldest carrier with a substantial number of planes.   



## Question 3:

* What states did Southwest Airlines' **flight paths** tend to fly to?
* What states did Southwest Airlines' **flights** tend to fly to?

For example, Southwest Airlines Flight 60 to Dallas consists of a single flight
path, but since it flew 299 times in 2013, it would be counted as 299 flights.

```{r, echo=FALSE, message=FALSE, fig.width=12, fig.height=6}

southwest <- flights %>%
  group_by(carrier) %>%
  filter(carrier=='WN')

airport_local <- left_join(airports, states, by="state")

southwest <- left_join(southwest, airport_local, by = c("dest"="iata")) %>%
  filter(cancelled==0) %>%
  select(state, flight, dest, plane, airport, city, date) %>%
  group_by(state)

totalflights <- southwest %>% 
  group_by(state) %>% 
  filter(!is.na(state)) %>%
  tally() %>% 
  rename(totalflights = n)


flightpaths <- southwest %>% 
  group_by(state, flight) %>% 
  summarize(length(state)) %>%
  filter(!is.na(state)) %>%
  tally() %>% 
  rename(flightpaths = n)



ggplot(totalflights, aes(x=fct_reorder(state, totalflights, median, na.rm=TRUE), y=totalflights)) +
  geom_bar(stat="identity", position="dodge", fill="violetred4") +
  labs(title="Flights by State", x="State", y="Number of Flights")

ggplot(flightpaths, aes(x=fct_reorder(state, flightpaths, median, na.rm=TRUE), y=flightpaths)) +
  geom_bar(stat="identity", position="dodge", fill="violetred4") +
  labs(title="Flightpaths by State", x="State", y="Number of Flightpaths")


```

**Interpretation** 
For both flights and flight paths, there were the most flights within Texas for flights to and from Houston.  The majority of the remaining flights and flightpaths are within the south, like Florida, Louisiana and Oklahoma.  California and Illinois also has a fair amount of flights to it.  I also removed cancelled flights from my analysis since those cancelled flights were never completed.    




## Question 4:

I want to know proportionately what regions (NE, south, west, midwest) each 
carrier flies to/from Houston in the month of July.  Consider the `month()`
function from the `lubridate` package.

```{r, echo=FALSE, message=FALSE, fig.width=12, fig.height=6}

airport_local <- left_join(airports, states, by="state")

houston <- left_join(airport_local, flights, by=c("iata" = "dest")) %>%
  filter(!is.na(plane) & cancelled==0) %>%
  select(carrier, iata, region, date) %>%
  group_by(carrier, region) %>%
  arrange(carrier) %>%
  mutate(month=month(date)) %>%
  filter(month==7)

carrier_totals <- houston %>% 
  group_by(carrier) %>% 
  tally() %>% 
  rename(carrier_totals = n)


carrier_region <- houston %>% 
  filter(!is.na(region)) %>%
  group_by(carrier, region) %>% 
  tally() %>% 
  rename(carrier_region = n)


carrier_region <- left_join(carrier_region, carrier_totals, by="carrier") %>%
  mutate(prop = (carrier_region/carrier_totals))
  
ggplot(carrier_region, aes(x=region, y=prop)) +
  geom_bar(stat="identity", position="dodge", fill="violetred4") +
  facet_wrap(~carrier) +
  labs(title="Proportion of Flights per Region by Carrier", x="Region", y="Proportion of Flights") 

```

**Interpretation** 
The majority of flights from Texas remain within the south.  As we saw earlier, the majority of flights are just within Texas to begin with.  For example, though American Airlines is an international carrier, its largest hub is in Dallas, so that is likely why American Airlines only flies within the south from Houston to Dallas.